---
title: "Data Transformation Exercise in R: Converting a nested list into dataframe"
author: Mert Nuhoglu
date: July 21, 2015
output: ioslides_presentation
---

```{r load_packages, include=FALSE}
library("dplyr")
library("data.table")
```

## Problem Definition

- Conversion of a list into a data frame (or data table)
- Lists: Hierarchical 
- Data Frames: Flat

<div class="notes">
A very common problem in data cleaning or data transformation jobs is the conversion of some list data structure into a data frame data structure. 

Lists in R are hierarchical data structures. Whereas data frames or data tables are flat data structures. This crucial difference makes the transformation of the data structures a bit cumbersome. @mp3=p001_01
</div>

## Source Data Structure

- To reproduce the list use the following code

```{r }
f = list(list(NA, "false", "5"), list(NA, "false", "3"))
```

<div class="notes">
In this example, we have a nested list of 2 levels. That is the list consists of elements which are themselves lists of some primitives.

Let's write this expression in R console and see the structure of the list f. @mp3=p002_01 @code
</div>

## Structure of List

```{r }
str(f)
```

<div class="notes">
Use str() function to see the structure of the list f. @code

This shows that f consists of a list of 2 elements. This is the upper or root level of our hierarchical data structure. 

Each element of these two lists consists of lists of three elements. Their elements are primitives. So there is no more nested lists. @mp3=p003_01
</div>

## Target Data Structure

- A dataframe of two rows and three columns:

			X1    X2 X3
	1 <NA> false  5
	2 <NA> false  3

<div class="notes">
Target structure is a data frame with 3 columns and 2 rows. Each list element should become a row.

So, the question is how can we transform the nested list f into this flat dataframe? @mp3=p004_01
</div>

## Alternative 1: Can we use rbind or rbindlist? 

```{r }
do.call(rbind, f)
```

```{r }
rbindlist(f)
```

<div class="notes">
In this case, we can use the function rbind or rbindlist because each element of the list f are lists that have the same structure. So each nested list corresponds to a row and we are actually rbinding these rows together. @mp3=p005_01

Note that there is a difference in the results of rbind and rbindlist. rbind returns a matrix whereas rbindlist returns a data.table. So, when we use rbind, we have to convert this matrix into a data.table with an additional operation. @mp3=p005_02
</div>

## Converting matrix to data.table

```{r }
do.call(rbind, f) %>%
	as.data.table
```

```{r }
do.call(rbind, f) %>%
	as.data.frame
```

<div class="notes">
To convert the matrix into a dataframe or datatable, we need to use as.data.table or as.data.frame functions. @mp3=p006_01
</div>

## List of dataframe

```{r }
g = list(
	data.frame(a=NA, b="false", c="5"), 
	data.frame(a=NA, b="false", c="3")
)
str(g)
```

<div class="notes">
Let's take another example. @mp3=p007_01

Note that, the elements of the list g are dataframes. There is no nested list structure in this object. g is list of two elements where each element is a dataframe. @mp3=p007_02

Whereas in the previous case, the elements of list f are again lists themselves. (@code compare f and g in console) @mp3=p007_03

</div>

## Can we use rbind or rbindlist? 2

```{r }
do.call(rbind, g)
```

```{r }
rbindlist(g)
```

<div class="notes">
Again, we can use the function rbind or rbindlist because each element of the list g is a dataframe. We are actually rbinding these dataframes together. @mp3=p008_01
</div>

## Does rbind/rbindlist work always?

```{r }
f2 = list(
	list(list(7, 2), "false", "5"), 
	list(list(1, 3), "false", "3")
)
```

```{r }
do.call(rbind, f2) %>%
	as.data.frame
```

```
rbindlist(f2)
```

	Error in rbindlist

<div class="notes">
Do rbind or rbindlist work always? Unfortunately, no. There are several cases of nested list structures where you have to find more complicated ways to transform them into dataframes. @mp3=p009_01

For example, take any list that has more than two levels of nested list. Calling rbindlist on them will result in error. Whereas calling rbind results in a dataframe that is a little weird.

Note that, the first column of the resulting data.frame is not a primitive.  @mp3=p009_02
</div>

## rbind with deeply nested list

```{r }
d2 = do.call(rbind, f2) %>%
	as.data.frame
str(d2)
```

<div class="notes">
Here we can see the first column of resulting data.frame more clearly. It is not a primitive data such as character or number. Instead it is a list data. So, the data.frame become a non-flat data structure. @mp3=p010_01

A non-flat data.frame is not something that R programmers expect to work with. It won't play nicely with subsetting functions. Therefore we should make additional operations to convert this data.frame into a proper flat data.frame. @mp3=p010_02
</div>

