---
title: "Guided Quiz: Sorting Data Frames and Tables By Columns"
author: Mert Nuhoglu
date: July 22, 2015
output: ioslides_presentation
---

```{r load_packages, include=FALSE}
library("dplyr")
```

# Introduction

<div class="notes">
Welcome. This is a new style for me to teach. I am trying to experiment different styles in order to get some insight on what works better.  @mp3=p001_01

I called this teaching style as guided quiz. Maybe there are other good names for it. I am not very sure because I am actually new to video training. But if there is a better and more common name for this teaching style, please let me know. @mp3=p001_02

So, what do I mean by guided quiz?

We all know what a quiz is. Usually, the quizes are done by the learner periodically while learning a new topic. It is very useful because it puts the learner into a more active mind state. @mp3=p001_03

I wonder if it is possible to join quizes with video training. What if there were some video trainings that would consist of quiz questions? Present a question to the learner. Let the learner think a bit. Then present the answer of the question with some explanation. Wouldn't it be useful? @mp3=p001_04

I will try to see if this works or not. Please let me know your own experience, so we can improve this style in a better direction. @mp3=p001_05
</div>

## How to sort a dataframe by columns

```{r }
	df = data.frame(
		name = c("Ali", "Olli", "Billy"),
		age = c(15, 30, 20),
		state = c("NY", "CA", "NY")
	)
```

<div class="notes">
So, what is the quiz problem that we are gonna solve now? @mp3=p002_01

Let's assume we have a dataframe as this. The dataframe has 3 columns: name, age, and state. @mp3=p002_02

We want to sort this dataframe by its columns. There are alternative ways to sort a dataframe. I will ask about some of these alternatives not all of them. @mp3=p002_03
</div>

### Sort by one column

- Sort the dataframe by its name column
- Ascending order

```{r }
	df = data.frame(
		name = c("Ali", "Olli", "Billy"),
		age = c(15, 30, 20),
		state = c("NY", "CA", "NY")
	)
	df
```

<div class="notes">
First question:

How do you sort this dataframe by its name column? The letter A should come first. That is the order should be ascending order. @mp3=p003_01
</div>

### Solution

```{r }
	df[ order(df$name), ]
```

<div class="notes">
Here is the answer.

You use the order function. The order function is inside the base or standard R library. The function expects a vector. The function orders this vector. The input argument can be of any primitive data type such as character, numeric or factor. The returned value is an integer vector that corresponds to the indices of the input vector. @mp3=p004_01
</div>

### order() function

```{r }
	order( c("Apple", "IBM", "Microsoft") )
```

```{r }
	order( c("IBM", "Microsoft", "Apple" ) )
```

<div class="notes">
For example, assume that we have a vector of three elements: "Apple", "IBM", "Microsoft". calling order() function on this vector returns an integer vector: 1, 2, 3. @mp3=p005_01

Now, let's change the order of the elements in the vector and call order() function again.  @mp3=p005_02

This time we obtained the result: 3, 1, 2. What do these numbers mean? @mp3=p005_03

3 is the first element in resulting vector. So this means that 3. element of input argument should be the first element when it is ordered. 1. element of input argument should be second element when it is ordered. @mp3=p005_04
</div>

### Solution

```{r }
	df[ order(df$name), ]
```

<div class="notes">
Now, return back to our solution. This solution consists of two parts: @mp3=p006_01

First, we have a dataframe object called as df

Second, we pass this dataframe object an expression called as: order(df$name) @mp3=p006_02

This expression returns an integer vector which is actually the order of indices of df$name column when they are in order. @mp3=p006_03

This expression is passed as the first argument to df object's bracket operator. Remember bracket operator takes two arguments. The first argument corresponds to the row indices, the second argument corresponds to column indices. @mp3=p006_04

Since, the expression order() is passed to the first argument, the result of it orders the rows of dataframe object according to the name column's order. @mp3=p006_05
</div>

### Question: Sort using datatable 

- Use datatable instead of dataframe

```{r }
	dt = data.table(df)
	dt
```

<div class="notes">
Next question: Sort the same way by name column but this time use a datatable instead of dataframe. @mp3=p007_01
</div>

### Solution

```{r }
	dt[ order(dt$name) ]
```

instead of 

	df[ order(df$name), ]

<div class="notes">
The solution is very similar to dataframe solution. This time we don't put an additional comma after the order() expression. @mp3=p008_01

Why don't we put comma? Because datatable's bracket operator assumes that the first argument is the indices of the rows whereas dataframe's bracket operator expects the indices of the columns first. @mp3=p008_02

I always prefer to use datatables instead of dataframes. Because indexing style is more intuitive and also it is much faster than dataframe. @mp3=p008_03
</div>

### Question: Reverse the order

- Descending order instead of ascending order
- That is: Letter Z should come first

<div class="notes">
Next question: Reverse the order.

Sort in the same way by using the name column of the data frame or table, but this time the letter Z should come first. Thus use descending order not ascending order. @mp3=p009_01
</div>

### Solution

```{r }
	df[ order(-df$name), ]
```

```{r }
	dt[ order(-dt$name), ]
```

<div class="notes">
This is very simple. Just put "minus" sign in front of the column that you want to be in descending order. @mp3=p010_01
</div>

### Solution 2

- Which one is first?

	5 10 20

- Now?

	-5 -10 -20

<div class="notes">
Of course, you might ask: Why did we put "minus" sign? What does it mean? How will I remember this rule now? @mp3=p011_01

To make remembering easier think about a very simple example. Assume that we have a vector of three elements: 5 10 20 @mp3=p011_02

Which one is first when ordered normally? Note that normal order is ascending order according to our common expectation. @mp3=p011_03

5 is first and 20 is last.

Now let's take the negatives of these numbers. Which one comes first when ordered naturally? @mp3=p011_04

The smallest one is the first one. And the smallest one is -20. So taking the negatives of the numbers reverses the order of them. That is a universal fact. It is always valid. And you can use it in order to remember the ordering rule better. @mp3=p011_05
</div>


### Question: Sort using dplyr style

- Sort using dplyr style
- Use pipes `%>%` and `arrange` functions

<div class="notes">
Next question: Sort using dplyr style. Namely use pipes and arrange function of dplyr. @mp3=p012_01

If you have not used dplyr till now, I highly recommend you to check tutorials on it. It is such a wonderful data handling tool. It is revolutionary because it made so many great improvements in the process of data cleaning and data analysis jobs. @mp3=p012_02

</div>

### Solution

```{r }
	dt %>%
		arrange( name )
```

```{r }
	df %>%
		arrange( name )
```

<div class="notes">
The solution with dplyr is much simpler than previous solutions.  @mp3=p013_01

We pass the data object dt or df as first argument using pipe operator. @mp3=p013_02

Then we pass the criteria of ordering as second argument to the arrange function. You might wonder why did I say second argument. There is just one argument inside the parantheses of arrange function. Where is the second argument? @mp3=p013_03
</div>

### Solution 2

- With pipes

```{r }
	dt %>%
		arrange( name )
```

- Without pipes

```{r }
	arrange( dt, name )
```

<div class="notes">
Pipe operator is a convenience function. It is optional. If we use common way of calling a function, then we can omit the pipes.  @mp3=p014_01

Here we pass to "arrange" function two arguments, as it is clear: dt and name expressions. In fact, pipe operator does exactly the same thing as this ordinary way of calling a function. It pipes the preceding argument as the first argument to the next function. @mp3=p014_02

Thus, pipe passes "dt" object as the first argument of "arrange" function. So, "name" becomes the second argument of arrange function. @mp3=p014_03
</div>

### Solution

```{r }
	dt %>%
		arrange( name )
```

```{r }
	df %>%
		arrange( name )
```

<div class="notes">
Back to our solution. 

Note that there is absolutely no difference between using datatable or dataframes. This is very cool because this kind of small differences cause lots of mistakes. @mp3=p015_01

I think that dplyr style is much more intuitive than using order function in base R library. Because of at least two reasons: @mp3=p015_02

1. You don't need to repeat the dataframe object every time in front of the column name.  @mp3=p015_03

2. All the dplyr data manipulation functions have the same basic interface. You don't need to remember all the differences between different functions. Always, the first argument is the data object. The second argument is the criteria of data manipulation. Additional arguments are usually specified with their parameter name. @mp3=p015_04

The intuitivity of the interface is very important for effective programming because it lets the programmer to get into the flow while working.  @mp3=p015_05
</div>


### Question: Using with and base order functions

- Use with() function
- Use base::order function
- Encapsulate the columns to order inside with

<div class="notes">
Next question. Using with and base order functions. @mp3=p016_01

There is a function called "with()" that lets the programmer to specify the column names without prepending them with their data containers. This might be used in order to make order() function to work a little bit like dplyr. How to use it to sort the dataframe? @mp3=p016_02
</div>

### Solution

```{r }
	df[ with(df, order(name)), ]
```

Compare the ordinary way:

```{r }
	df[ order(-df$name), ]
```

<div class="notes">
Here is the solution. In this case, it is not actually easier to write nor read than ordinary way. @mp3=p017_01

You have to repeat the data container object inside with. But you don't need to repeat it again inside order expression because order expression is passed as an argument to with() function. @mp3=p017_02

Putting a new function like with() increases the burden on the programmer because now we need to check the order of the parameters and parantheses. This is more difficult to write and read.  @mp3=p017_03
</div>

### Reverse the order in dplyr style

Sort using dplyr style. Reverse the order.

<div class="notes">
Next question: Reverse the order. Use dplyr style. @mp3=p018_01

Remember in base order() function, we need to put "minus" operator in front of the column expressions that should be in descending order. How to reverse the order in dplyr? @mp3=p018_02
</div>

### Solution

```{r }
	dt %>%
		arrange( desc(name) )
```

<div class="notes">
In this case, we need to wrap the column "name" inside the "desc" function. The name of this function is the abbreviation for the word "descending". @mp3=p019_01

Not a big diffence from ordinary way of putting "minus" operator in front. But still I think dplyr style is easier to read because you don't need to remember that "minus" corresponds to "descending" order but not to "ascending" order. @mp3=p019_02
</div>

## Sort by two columns

- Sort by two columns: state (ascending) and age (descending)
- Use dataframe and base order()

<div class="notes">
Next question: Sort by two columns

First, sort using dataframes and base order function. But this time sort by two columns not one column. First order in ascending order of state name, then sort by descending order of age. @mp3=p020_01
</div>

## Solution

```{r }
	df[ order(df$state, -df$age), ]
```

<div class="notes">
Here is the solution. This time we pass multiple arguments to order() function.

Note that, the df$age expression has a "minus" operator in front. @mp3=p021_01
</div>

## Use datatable

- Same question
- Use datatable instead of dataframe
- Sort by two columns: state (ascending) and age (descending)

<div class="notes">
Now, do the same sorting but use datatable instead of dataframes.  @mp3=p022_01
</div>

## Solution

- No comma sign

```{r }
	dt[ order(dt$state, -df$age) ]
```

<div class="notes">
Again a very similar solution as before. The only difference is that datatable does not have a comma after order() expression. @mp3=p023_01
</div>

### Use dplyr functions

- Same question
- Use dplyr style
- Sort by two columns: state (ascending) and age (descending)

<div class="notes">
Next question: Do the same sorting but this time use dplyr functions: pipe and arrange. @mp3=p024_01
</div>

## Solution

```{r }
	dt %>%
		arrange( state, desc(age) )
```

<div class="notes">
Solution is here. We pass two arguments to arrange() function. First argument is "state" column. Second argument is the expression "desc(age)".

The code is still pretty clear to read and write. @mp3=p025_01
</div>

### Conclusion

- Which style is better?
- Personal tastes
- My choice: dplyr
- Declarative programming style:
	- What to do?
	- Not how to do it?

<div class="notes">
These are just a small set of alternatives to sort data frames and data tables. There are much more alternatives to do this. But I think these are the most common ones. @mp3=p026_01

So which one is better? Should we learn and use all of them? Or should we stick to one style? If so, which one? @mp3=p026_02

Everybody has a different taste when it comes to grammar rules. My choice of interface is dplyr. I think that dplyr is the most intuitive style among the alternatives.  @mp3=p026_03

The fundamental reason for this is that dplyr has a declarative programming style. That is it lets the programmer think in terms of "what needs to be done" not in terms of "how to do something".  @mp3=p026_04

Even in this small example, using datatable or dataframe with base order function requires the programmer to think in implementation details. The programmer needs to think about the order of indexes first and then combine that with the ordering of datatable. This puts an additional mental burden on our minds. dplyr abstracts the programmer from this kind implementation details completely.  @mp3=p026_05
</div>



